Characteristic polynomials of p-adic matrices

### 2017-01-30 charpoly/charpoly.tex
Short form of Com(X-M)

In light of Proposition \ref{prop:mainlem}, we recall the computation of $d\chi$
\citelist{\cite{caruso-roe-vaccon:14a}*{Appendix B} \cite{caruso-roe-vaccon:15a}*{\S 3.3}}.
Let $\com(M)$ denote the comatrix of $M$; when $M \in \GL_n(K)$ then $\com(M) = \det(M) M^{-1}$.  Then $d\chi(M)$ is given
by
\[
d\chi(M): dM \mapsto \tr(\com(X-M) \cdot dM).
\]
Moreover, $d\chi(M)$ will be surjective exactly when the Jordan blocks of $M$ all have distinct generalized eigenvalues.

\todo{Formula with comatrix.}

\subsection{Properties of the image of $d \chi$}

\todo{Surjectivity. Stability under multiplication by $X$ mod $\chi$.}

@@

\begin{prop} \label{prop:factor_com}
Let $M \in M_n (O_K)$ be such that $discr(\chi (M)) \neq 0.$
We assume that $A=\com (X-M)$ is such that $gcd(A_{1,1}, \chi (M))=1$
(in $K[X]$).
Then for $U=col(A,1)$ and $V=A_{1,1}^{-1} row(A,1) \mod \chi (M),$
both in $O_K[X]^n,$ we have: 
\[ \com (X-M)= U \cdot {}^t V \mod \chi (M).\]
\end{prop}
\begin{proof}
We first prove that for all $i,j \in \llbracket 1, n \rrbracket,$ we have
\begin{equation}A_{1,1} A_{i,j}=A_{1,i}A_{1,j} \mod \chi (M). \label{eqn:2-minors-com} \end{equation}
Let $\lambda$ be a root of $\chi (M)$ in an algebraic closure of $K.$
Then $A(\lambda)$ is of rank one.
Indeed, $A(\lambda) \times (\lambda-M) =0.$
Since $\chi (M)$ is square-free, then $\lambda-M$ is of rank $n-1$
and then $A(\lambda)$ is of rank at most one.
Nevertheless, we can remark that
 $gcd(A_{1,1}, \chi (M))=1,$ $A_{1,1} \neq 0 \mod (X-\lambda)$
and thus, $A(\lambda)$ is of rank exactly one.
As a consequence, all $2 \times 2$-minors of $A(\lambda)$ are zero.
It means that \[A_{1,1} A_{i,j}=A_{1,i}A_{1,j} \mod (X-\lambda). \]
By the Chinese Remainder Theorem, our first result is then proved.

@@

\begin{rem}
If $M \in M_n (O_K)$ is such that for some $U,V \in O_{K}[X]^n$
we have $\com (X-M)= U \cdot {}^t V \mod \chi (M)$ and if 
$P\in GL_n (O_K)$ then
\[\com (X-PMP^{-1})=(PU) \cdot ( {}^tV P^{-1}) \mod \chi(M).\]
In other words, factorizability of $\com(X-M)$ is stable
by change of basis.
\end{rem}

\section{Optimal precision\\via Hessenberg form}
%\section{Computation of the characteristic polynomial and its derivative}

@@

\section{Optimal precision\\via Frobenius form}

\section{Eigenvalues}

>>>>

In light of Proposition \ref{prop:mainlem}, we recall the computation of $d\chi$
\citelist{\cite{caruso-roe-vaccon:14a}*{Appendix B} \cite{caruso-roe-vaccon:15a}*{\S 3.3}}.
Let $\com(M)$ denote the comatrix of $M$; when $M \in \GL_n(K)$ then $\com(M) = \det(M) M^{-1}$.  Then $d\chi_M$ is given
by
\[
d\chi_M: dM \mapsto \tr(\com(X-M) \cdot dM).
\]
%Moreover, $d\chi(M)$ will be surjective exactly when

%\subsection{Properties of the image of $d \chi$}

\begin{prop}
\label{prop:surjectivity}
For $M \in M_n(K)$, the following conditions are equivalent:
\begin{enumerate}[(i)]
\renewcommand{\itemsep}{0pt}
\item the differential $d\chi(M)$ is surjective
\item the matrix $M$ has a cyclic vector (\emph{i.e.} $M$ is similar
to a companion matrix)
\item the eigenspaces of $M$ associated to eigenvalues in an algebraic
closure of $\bar K$ have all dimension $1$
%\item the Jordan blocks of $M$ all have distinct generalized eigenvalues
\end{enumerate}
\end{prop}

By definition, the codomain of $d \chi_M$ is $K_{< n}[X]$. 
However, when $M$ is given, $K_{< n}[X]$ is canonically isomorphic
to $K[X]/\chi_M(X)$ as a $K$-vector space. For our purpose, it will 
often be convenient to view $d \chi_M$ as an $K$-linear mapping
$M_n(K) \to K[X]/\chi_M(X)$.

\begin{prop}
Let $A$ be the subring of $K[X]$ consisting of polynomials $P$ for
which $P(M) \in M_n(\OK)$. Then $d \chi_M \big(M_n(\OK)\big)$ 
(considered as a submodule of $K[X]/\chi_M(X)$) is  stable by 
multiplication by the elements of $A$.
\end{prop}

\begin{proof}
\todo{The image is generated by the entries of $C = \com(X-M)$
and we have $P(M) \cdot C \equiv  P(X) \cdot C \pmod{\chi_M(X)}$.}
\end{proof}

\begin{cor}
If the coefficients of $M$ lie in $\OK$, then $d \chi_M \big(M_n(\OK)\big)$ 
is stable by multiplication by $X$ and hence is a module over $\OK[X]$.
\end{cor}

@@

\begin{comment}
\begin{prop} \label{prop:factor_com}
Let $M \in M_n (O_K)$ be such that $discr(\chi (M)) \neq 0.$
We assume that $A=\com (X-M)$ is such that $gcd(A_{1,1}, \chi (M))=1$
(in $K[X]$).
Then for $U=col(A,1)$ and $V=A_{1,1}^{-1} row(A,1) \mod \chi (M),$
both in $O_K[X]^n,$ we have: 
\[ \com (X-M)= U \cdot {}^t V \mod \chi (M).\]
\todo{I think that the correct condition is $d \chi_M$ surjective (which
is a bit weaker).}
\end{prop}

\begin{proof}
We first prove that for all $i,j \in \llbracket 1, n \rrbracket,$ we have
\begin{equation}A_{1,1} A_{i,j}=A_{1,i}A_{1,j} \mod \chi (M). \label{eqn:2-minors-com} \end{equation}
Let $\lambda$ be a root of $\chi (M)$ in an algebraic closure of $K.$
Then $A(\lambda)$ is of rank one.
Indeed, $A(\lambda) \times (\lambda-M) =0.$
Since $\chi (M)$ is square-free, then $\lambda-M$ is of rank $n-1$
and then $A(\lambda)$ is of rank at most one.
Nevertheless, we can remark that
 $gcd(A_{1,1}, \chi (M))=1,$ $A_{1,1} \neq 0 \mod (X-\lambda)$
and thus, $A(\lambda)$ is of rank exactly one.
As a consequence, all $2 \times 2$-minors of $A(\lambda)$ are zero.
It means that \[A_{1,1} A_{i,j}=A_{1,i}A_{1,j} \mod (X-\lambda). \]
By the Chinese Remainder Theorem, our first result is then proved.

@@

\begin{rem}
If $M \in M_n (O_K)$ is such that for some $U,V \in O_{K}[X]^n$
we have $\com (X-M)= U \cdot {}^t V \mod \chi (M)$ and if 
$P\in GL_n (O_K)$ then
\[\com (X-PMP^{-1})=(PU) \cdot ( {}^tV P^{-1}) \mod \chi(M).\]
In other words, factorizability of $\com(X-M)$ is stable
by change of basis.
\end{rem}
\end{comment}

\section{Differential\\via Hessenberg form}
%\section{Computation of the characteristic polynomial and its derivative}

@@

\section{Differential\\via Frobenius form}

The algorithm designed in the previous section computes the differential 
$d \chi_M$ of $\chi$ at a given matrix $M \in M_n(K)$ for a cost of 
$O(n^3)$ operations in $K$. This seems to be optimal given that the 
(naive) size of the $d \chi_M$ is $n^3$: it is a matrix of size $n
\times n^2$. It turns out however that improvements are still possible!

Precisely, we will prove in this section that $d \chi_M$ has a very 
special form, allowing a compact encoding using only $O(n^2)$ coefficients. 
We shall moreover design fast algorithms (with complexity $\softO
(n^\omega)$) for computing it. The price to pay is that divisions in
$K$ appear, which can be an issue regarding to precision in particular
cases.

From now on, we fix a matrix $M \in M_n(K)$ for which $d \chi_M$ is 
surjective.

\subsection{Compact form of $d \chi_M$}

Let $C$ be the companion matrix associated to $\chi_M$:
$$N = \left( \begin{matrix}
0 & 1 & 0 & \cdots & 0 \\
\vdots & \ddots & 1 & \ddots & \vdots \\
\vdots & & \ddots & \ddots & 0 \\
0 & \cdots & \cdots & 0 & 1 \\
-a_0 & -a_1 & \cdots & \cdots & -a_{n-1}
\end{matrix} \right)$$
with $\chi_M = a_0 + a_1 X + \cdots + a_{n-1} X^{n-1} + X^n$.
By Proposition~\ref{prop:surjectivity}, there exists a matrix 
$P \in \GL_n(K)$ such that $M = P N P^{-1}$. Applying the same
result to the transpose of $M$, we find that there exists another
invertible matrix $Q \in \GL_n(K)$ such that $M^t = Q N Q^{-1}$.

\begin{prop}
\label{prop:shortcom}
We keep the previous notations and assumptions.
Let $V$ be the row vector $(1, X, \ldots, X^{n-1})$. Then
$$\com(X{-}M) = \alpha \cdot P V^t \cdot V Q^t
\mod \chi_M$$
for some $\alpha \in K[X]$.
\end{prop}

\begin{proof}
Write $C = \com(X{-}M)$. From $(X{-}M) \cdot C \equiv 0 
\pmod{\chi_M}$, we deduce $(X{-}N) \cdot P^{-1} C \equiv 0 \pmod{\chi_M}$. 
Therefore each column of $P^{-1} C$ lies in the right kernel of $X{-}N$
modulo $\chi_M$. On the other hand, a direct computation shows that
every column vector $W$ lying in the right kernel of $X{-}N$ modulo 
$\chi_M$ can be written as $W = w \cdot V^t$ for some $w \in 
K[X]/\chi_M$. We deduce that $C \equiv P \cdot V^t B \pmod{\chi_M}$
for some row vector $B$.
Applying the same reasoning with $M^t$, we find that $B$ can be
written $B = \alpha V Q^t$ for some $\alpha \in K[X]/\chi_M$ and
we are done.
\end{proof}

Proposition~\ref{prop:shortcom} shows that $\com(X{-}M)$ can be encoded 
by the datum of the quadruple $(\alpha, P, Q, \chi_M)$ whose total size 
stays within $O(n^2)$ : the polynomials $\alpha$ and $\chi_M$ are 
determined by $2n$ coefficients while we need $2n^2$ coefficients to 
write down the matrices $P$ and $Q$. 
We shall see moreover in \S \ref{sec:eigenvalues} that interesting
informations can be read off on this short form $(\alpha, P, Q, 
\chi_M)$.

\subsection{Fast computation}

\subsection{Optimal jagged precision}

\section{Eigenvalues}
\label{sec:eigenvalues}
